BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-09 implementation.tex
First Batch of fixes (Emilio's)

\chapter{Architecture and Implementation}
In this chapter we discuss the implementation of BlueTracer, providing a detailed account on how the tool is organized, on the design choices which were made during its development process and on how encountered challenges were dealt with.

@@

\begin{itemize}
\item \textbf{User-friendliness}. Pin's user model enables the insertion of analysis code in arbitrary locations of the binary through the use of its uncomplicated but rich C and C++ APIs.
\item \textbf{Portability}. Pin's APIs are independent from the architecture but, at the same time, they allow the extraction of architecture-specific information.
\item \textbf{Transparency}. The information obtained by Pin perfectly describes the program's original actions. This because an application instrumented by Pin preserves instruction and data addresses as well as memory and register values of the uninstrumented execution. 
\item \textbf{Efficiency}. Pin employs a just-in-time (JIT) in order to add and optimize analysis code. It makes use of a set of optimization techniques, including code caching and inlining.
\item \textbf{Robustness}. Pin uncovers code at run-time and is thus able to manage dynamically generated code, dynamically loaded libraries and the targets of indirect jumps which are statically unknown. 
\end{itemize}

@@

In light of what just stated, BlueTracer's integration with BluePill is straightforward, since it essentially complements the \textit{Target platform simulation layer}, as it can be observed from \textit{Figure 3.1}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.65\textwidth]{Figures/BluePill.pdf}
\caption{\textit{BlueTracer's integration with BluePill}}
\end{figure}

@@

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{Native APIs tracing}, the second for \textbf{APIs tracing} and the last focused on \textbf{user-mode callbacks and Windows APCs tracing} (\textit{Figure 3.2}). In particular, the tracing of Native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{PyREbox}'s database).

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\end{figure}

@@

Pin provides a set of APIs aimed at assisting the extraction of information relative to the system calls made by the pinned application, also including information relative to Windows Native APIs. In particular, in BluePill (and consequently in BlueTracer) \texttt{PIN\_AddSyscallEntryFunction} and  \texttt{PIN\_AddSyscallExitFunction} were used for this purpose. In fact, these allow to register notification functions which are called immediately before and after the execution of a system call \cite{Pin}. In BlueTracer, the function in charge of gathering Native API information before execution is \texttt{TraceSysCallEntry} while the one responsible for collecting Native API information after execution is \texttt{TraceSysCallExit}. They have been both been embedded in BluePill's notification functions and their overall structure is detailed in \textit{Figure 3.3}.
\\
\begin{figure}[h]
\centering
\includegraphics[scale=0.75]{Figures/NTAPITracing.pdf}
\caption{\textit{Native APIs Tracing Workflow}}
\end{figure}
\\
Following \textit{Figure 3.3}, let us now thoroughly analyze the main steps which take place when Native APIs are traced.

\subsection{Main Image Check}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether or not only Native APIs called invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure 3.3}, the first thing which is done in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the state depicted by \textit{Figure 3.4}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}     
Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

\begin{enumerate}
\item The first check involves discovering whether or not \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item Secondly, it is required to examine if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure 3.4}, such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether or not all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{structs} (\textit{Listing 3.2}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e. 18), since a Native API can have at most 18 arguments. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure 3.3}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building an hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Native API Logging}
With reference to \textit{Figure 3.3}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamic-link libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g. file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a very high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure 3.5}.

Now let us illustrate the actions being performed during API tracing by analyzing in detail each one of the three main blocks of \textit{Figure 3.5}.

@@

Similarly to what occurs for NTAPI tracing, in order to log as much information as possible and to correctly record the arguments' values, it is necessary to have a source of API information, providing some kind of assistance during the tracing activity. When dealing with APIs, BlueTracer allows the user to choose between two possible sources of API information: \textit{drltrace}'s configuration file or the data extracted from \textit{PyREbox}'s database. By default, the latter is employed as it embodies a greater amount of information, including variable names and named constants' enumerations.

@@

Each piece of information in an entry is separated by \texttt{|}. The first bit of data is the API's return type, the second is the API's name and the rest of the entry is made up by the arguments' types. In particular, the \texttt{\_\_out} token is used to mark output arguments, while \texttt{\_\_inout} is utilized to identify input and output arguments. In case there is no token, it means that the argument is just an input one.

@@

\newpage
To give a better idea on how API data is represented, an instance of \texttt{libcall\_arg\_info\_t} containing the information associated to \texttt{WriteFileEx} can be observed below (\textit{Listing 3.12}).
\\
\begin{lstlisting}[caption={Instance of \texttt{libcall\_arg\_info\_t} relative to \texttt{WriteFileEx}},captionpos=b]
	{ "WriteFileEx",5,
	{ 
		{-1, "Return value", NKT_DBFUNDTYPE_SignedDoubleWord, 0, 4, INOUT },
		{0, "hFile", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{1, "lpBuffer", NKT_DBFUNDTYPE_Void | NKT_DBOBJCLASS_Pointer, 0, 0, IN },
		{2, "nNumberOfBytesToWrite", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{3, "lpOverlapped", NKT_DBOBJCLASS_Struct | NKT_DBOBJCLASS_Pointer,
		 "_OVERLAPPED", 160, INOUT },
		{4, "lpCompletionRoutine", NKT_DBOBJCLASS_Typedef, 0, 0, IN },
	}
	}
\end{lstlisting}

Due to the high amount of APIs for which information was stored in the \textit{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure 3.6}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\end{figure}

Let us conclude this section by explaining how analysis routines are inserted before and after the API execution during the API instrumentation procedure outlined in pages 34-35. In BlueTracer such operation heavily relies on the use of the \texttt{RTN\_InsertCall}, which allows to insert analysis functions relative to a \texttt{RTN} object. 
\newpage
This API takes as input the following parameters:
\begin{itemize}
\item The routine object (\texttt{RTN}) representing the function under analysis.
\item A member of the \texttt{IPOINT} enumeration which determines where the analysis function is inserted in relation to the instrumented object. It can be either \texttt{IPOINT\_BEFORE}, in which case the analysis code is placed before the instrumented function, or \texttt{IPOINT\_AFTER}, in which case the analysis code is placed after the instrumented function.
\item The analysis function to be called.
\item The arguments to pass to the analysis function, specified via members of the \texttt{IARG\_TYPE} enumeration. If the analysis function is placed before the instrumented object, then such arguments may also include the parameters' values of the function being instrumented. However, for this to occur the pair <\texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE}, argument number> needs to be included for each of the analyzed function's parameters. Conversely, to obtain the instrumented function's result, \texttt{IARG\_FUNCRET\_EXITPOINT\_VALUE} is required. Obviously, this is possible only if the analysis function is placed after the instrumented object.
\end{itemize}

@@

In this section we will describe how APIs are analyzed before execution, i.e. the main steps that are carried out in the \texttt{PrintInfoBefore} functions, an overview of which is given in \textit{Figure 3.5}.
The workflow in this case is mostly the same as the one for the analysis of Native APIs, with one major difference, that is the introduction of a shadow stack used to store API entries in the thread local storage, which will be motivated later in the section.

@@

} apicall_t;
\end{lstlisting}
\vspace{1cm}
The \texttt{apicall\_t struct} is initialized utilizing the API-related data, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in section 3.4.1.

@@

Therefore, in light of what just stated, the last operations which are done when tracing an API before its execution are allocating a shadow stack entry (\textit{Listing 3.15}), initializing it and placing it on top of the shadow stack in the thread local data, so that it can be retrieved after the API's execution.
\vspace{3cm}
\begin{lstlisting}[caption={\texttt{struct} representing a shadow stack entry},captionpos=b]
typedef struct _stackEntry {
	apicall_t* apiInfo;    // API Call Reference
	ADDRINT  currentSP;    // Stack Pointer Register Content
	THREADID threadID;     // Thread ID
	uint api_counter;	   // Call Counter
} stackEntry;
\end{lstlisting}

@@

Lastly, let us detail the most important operations performed towards tracing API-related information after the API execution. These are carried out in the \texttt{PrintInfoAfter} function and can be observed from \textit{Figure 3.5}.

@@

In light of what just stated, it is natural for BlueTracer's logging activity, taking place in the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function, to involve the record of a callback occurrence, together with the name of the callback that is about to be invoked. Unfortunately, tracing the callback's arguments would require further work since they are packed in a data structure, as previously explained.
\newpage
Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.7}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

@@

\begin{enumerate}
\item The stack pointer value is retrieved. This is achieved through the use of the \texttt{PIN\_GetContextReg} API, that allows to get the value of a register in the context provided as input. Such context, in our specific case, is the application's register state after the context change, which, as previously mentioned, is passed as argument to the \texttt{CONTEXT\_CHANGE\_CALLBACK} notification function.
\item The callback number is obtained by adding 4 to the stack pointer and fetching the pointed value, in accordance to \textit{Figure 3.7}.
\item The \texttt{KernelCallbackTable} is retrieved from the PEB. 
\item The function pointer stored in the \texttt{KernelCallbackTable} entry indicated by the callback number is acquired.
\item It is checked if the address of the function falls within \texttt{User32.dll} memory address range. In fact, in \texttt{Win32} user mode callbacks are employed exclusively by \texttt{User32.dll} for windowing related aspects \cite{UCDispatcher}, meaning that the retrieved function address must belong to \texttt{User32.dll}'s memory address range. Such range is determined by retrieving an handle for \texttt{User32.dll} through \texttt{GetModuleHandle}, building an \texttt{IMG} object from it utilizing Pin's \texttt{IMG\_FindByAddress} and invoking the usual \texttt{IMG\_LowAddress} and \texttt{IMG\_HighAddress} functions. 
\item The name of the callback is determined by creating a \texttt{RTN} object from the function address via \texttt{RTN\_FindByAddress} and consequently invoking \texttt{RTN\_Name} on such object.
\item The callback occurrence and the callback name is recorded in the log.
\end{enumerate}

>>>>

\chapter{Architecture and Implementation} \label{implementation}
In this chapter we discuss the implementation of BlueTracer, providing a detailed account on how the tool is organized, on the design choices which were made during its development process and on how encountered challenges were dealt with.

It is important to highlight the fact that the material presented in this chapter is very dense and technical. We feel like there are no existing works which thoroughly explain the underlying technical details of a tool like BlueTracer and, in light of this, we think that giving a detailed account of the inner workings of such a tool would greatly benefit the community.

@@

\begin{itemize}
\item \textbf{User-friendliness}. Pin's user model enables the insertion of analysis code in arbitrary locations of the binary through the use of its uncomplicated but rich C and C++ APIs.
\item \textbf{Portability}. Pin's APIs are independent from the architecture but, at the same time, they allow the extraction of architecture-specific information.
\item \textbf{Transparency}. The information obtained by Pin perfectly describes the program's original actions. This because an application instrumented by Pin preserves instruction and data addresses as well as memory and register values of the uninstrumented execution. 
\item \textbf{Efficiency}. Pin employs a just-in-time (JIT) compiler in order to add and optimize analysis code. It makes use of a set of optimization techniques, including code caching and inlining.
\item \textbf{Robustness}. Pin uncovers code at run-time and is thus able to manage dynamically generated code, dynamically loaded libraries and the targets of indirect jumps which are statically unknown. 
\end{itemize}

@@

In light of what just stated, BlueTracer's integration with BluePill is straightforward, since it essentially complements the \textit{Target platform simulation layer}, as it can be observed from \textit{Figure~\ref{BluePill}}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.65\textwidth]{Figures/BluePill.pdf}
\caption{\textit{BlueTracer's integration with BluePill}}
\label{BluePill}
\end{figure}

@@

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{Native APIs tracing}, i.e., the tracing of the lowest-level user mode APIs,  the second for \textbf{APIs tracing} and the last focused on \textbf{user-mode callbacks and Windows APCs tracing} (\textit{Figure~\ref{structure}}). In particular, the tracing of Native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{PyREbox}'s database).

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\label{structure}
\end{figure}

@@

Pin provides a set of APIs aimed at assisting the extraction of information relative to the system calls made by the pinned application, also including information relative to Windows Native APIs. In particular, in BluePill (and consequently in BlueTracer) \texttt{PIN\_AddSyscallEntryFunction} and  \texttt{PIN\_AddSyscallExitFunction} were used for this purpose. In fact, these allow to register notification functions which are called immediately before and after the execution of a system call \cite{Pin}. In BlueTracer, the function in charge of gathering Native API information before execution is \texttt{TraceSysCallEntry} while the one responsible for collecting Native API information after execution is \texttt{TraceSysCallExit}. They have been both been embedded in BluePill's notification functions and their overall structure is detailed in \textit{Figure~\ref{NTAPI Tracing}}.
\\
\begin{figure}[h]
\centering
\includegraphics[scale=0.75]{Figures/NTAPITracing.pdf}
\caption{\textit{Native APIs Tracing Workflow}}
\label{NTAPI Tracing}
\end{figure}
\\
Following \textit{Figure~\ref{NTAPI Tracing}}, let us now thoroughly analyze the main steps which take place when Native APIs are traced.

\subsection{Main Image Check}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether or not only Native APIs called invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure~\ref{NTAPI Tracing}}, the first thing which is done in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the state depicted by \textit{Figure~\ref{sysenter}}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\label{sysenter}
\end{figure}     
Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

\begin{enumerate}
\item The first check involves discovering whether or not \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item If the first check fails, the second one is carried out, in which it is examined if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure~\ref{sysenter}}, such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

@@

For each Native API, the information related to it is contained in a \texttt{struct} of type \texttt{syscall\_info\_t} (\textit{Listing 3.2}). Most fields of such \texttt{struct} are self-explanatory. In particular, \texttt{num} is a \texttt{struct} storing two values indicating the system call number; these are filled in dynamically by Dr.Memory but are not needed by BlueTracer. Furthermore, the \texttt{flags} field is utilized to notify whether or not all the details of the Native API are known, as most Native APIs are undocumented. Undoubtedly though, a major role is played by the \texttt{arg} array, which is made up by \texttt{sysinfo\_arg\_t} \texttt{struct}s (\textit{Listing 3.3}) containing the data of the Native API arguments. This array is initialized with size \texttt{MAX\_ARGS\_IN\_ENTRY} (i.e. 18), since a Native API can have at most 18 arguments. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing Native API-related information},captionpos=b]
	typedef struct _syscall_info_t {

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure~\ref{NTAPI Tracing}}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building an hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Native API Logging}
With reference to \textit{Figure~\ref{NTAPI Tracing}}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamic-link libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g. file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a very high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\label{API Tracing}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure~\ref{API Tracing}}.

Now let us illustrate the actions being performed during API tracing by analyzing in detail each one of the three main blocks of \textit{Figure~\ref{API Tracing}}.

@@

Similarly to what occurs for Native API tracing, in order to log as much information as possible and to correctly record the arguments values, it is necessary to have a source of API information, providing some kind of assistance during the tracing activity. When dealing with APIs, BlueTracer allows the user to choose between two possible sources of API information: \textit{drltrace}'s configuration file or the data extracted from \textit{PyREbox}'s database. By default, the latter is employed as it embodies a greater amount of information, including variable names and named constants' enumerations.

@@

Each piece of information in an entry is separated by \texttt{|}. The first bit of data is the API's return type, the second is the API's name and the rest of the entry is made up by the arguments' types. In particular, the \texttt{\_\_out} token is used to mark output arguments, while \texttt{\_\_inout} is utilized to identify input and output arguments. In case there is no token, it means that the argument is just an input one. Essentially, an output parameter is a non-\texttt{null} pointer pointing to an element buffer, where the invoked function initializes the element.

@@

To give a better idea on how API data is represented, an instance of \texttt{libcall\_arg\_info\_t} containing the information associated to \texttt{WriteFileEx} can be observed below (\textit{Listing 3.12}).
\\
\begin{lstlisting}[caption={Instance of \texttt{libcall\_arg\_info\_t} relative to \texttt{WriteFileEx}},captionpos=b]
	{ "WriteFileEx",5,
	{ 
		{-1, "Return value", NKT_DBFUNDTYPE_SignedDoubleWord, 0, 4, INOUT },
		{0, "hFile", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{1, "lpBuffer", NKT_DBFUNDTYPE_Void | NKT_DBOBJCLASS_Pointer, 0, 0, IN },
		{2, "nNumberOfBytesToWrite", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{3, "lpOverlapped", NKT_DBOBJCLASS_Struct | NKT_DBOBJCLASS_Pointer,
		 "_OVERLAPPED", 160, INOUT },
		{4, "lpCompletionRoutine", NKT_DBOBJCLASS_Typedef, 0, 0, IN },
	}
	}
\end{lstlisting}

Due to the high amount of APIs for which information was stored in the \textit{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure~\ref{Hashmap}}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\label{Hashmap}
\end{figure}

Let us conclude this section by explaining how analysis routines are inserted before and after the API execution during the API instrumentation procedure outlined in pages 34-35. In BlueTracer such operation heavily relies on the use of the \texttt{RTN\_InsertCall}, which allows to insert analysis functions relative to a \texttt{RTN} object.

This API takes as input the following parameters:
\begin{itemize}
\item The routine object (\texttt{RTN}) representing the function under analysis.
\item A member of the \texttt{IPOINT} enumeration which determines where the analysis function is inserted in relation to the instrumented object. It can be either \texttt{IPOINT\_BEFORE}, in which case the analysis code is placed before the instrumented function, or \texttt{IPOINT\_AFTER}, in which case the analysis code is placed after the instrumented function.
\item The analysis function to be called.
\item The arguments to pass to the analysis function, specified via members of the \texttt{IARG\_TYPE} enumeration. If the analysis function is placed before the instrumented object, then such arguments may also include the parameters' values of the function being instrumented. However, for this to occur the pair <\texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE}, argument number> needs to be included for each of the analyzed function's parameters. Conversely, to obtain the instrumented function's result, \texttt{IARG\_FUNCRET\_EXITPOINT\_VALUE} is required. Obviously, this is possible only if the analysis function is placed after the instrumented object.
\end{itemize}

@@

In this section we will describe how APIs are analyzed before execution, i.e. the main steps that are carried out in the \texttt{PrintInfoBefore} functions, an overview of which is given in \textit{Figure~\ref{API Tracing}}.
The workflow in this case is mostly the same as the one for the analysis of Native APIs, with one major difference, that is the introduction of a shadow stack used to store API entries in the thread local storage, which will be motivated later in the section.

@@

} apicall_t;
\end{lstlisting}

The \texttt{apicall\_t struct} is initialized utilizing the API-related data, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in section 3.4.1.

@@

Therefore, in light of what just stated, the last operations which are done when tracing an API before its execution are allocating a shadow stack entry (\textit{Listing 3.15}), initializing it and placing it on top of the shadow stack in the thread local data, so that it can be retrieved after the API's execution.
\vspace{0.5cm}
\begin{lstlisting}[caption={\texttt{struct} representing a shadow stack entry},captionpos=b]
typedef struct _stackEntry {
	apicall_t* apiInfo;    // API Call Reference
	ADDRINT  currentSP;    // Stack Pointer Register Content
	THREADID threadID;     // Thread ID
	uint api_counter;	   // Call Counter
} stackEntry;
\end{lstlisting}

@@

Lastly, let us detail the most important operations performed towards tracing API-related information after the API execution. These are carried out in the \texttt{PrintInfoAfter} function and can be observed from \textit{Figure~\ref{API Tracing}}.

@@

In light of what just stated, it is natural for BlueTracer's logging activity, taking place in the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function, to involve the record of a callback occurrence, together with the name of the callback that is about to be invoked. Unfortunately, tracing the callback's arguments would require further work since they are packed in a data structure, as previously explained.

Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure~\ref{KiUserCallback}}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\label{KiUserCallback}
\end{figure}

@@

\begin{enumerate}
\item The stack pointer value is retrieved. This is achieved through the use of the \texttt{PIN\_GetContextReg} API, that allows to get the value of a register in the context provided as input. Such context, in our specific case, is the application's register state after the context change, which, as previously mentioned, is passed as argument to the \texttt{CONTEXT\_CHANGE\_CALLBACK} notification function.
\item The callback number is obtained by adding 4 to the stack pointer and fetching the pointed value, in accordance to \textit{Figure~\ref{KiUserCallback}}.
\item The \texttt{KernelCallbackTable} is retrieved from the PEB. 
\item The function pointer stored in the \texttt{KernelCallbackTable} entry indicated by the callback number is acquired.
\item It is checked if the address of the function falls within \texttt{User32.dll} memory address range. In fact, in \texttt{Win32} user mode callbacks are employed exclusively by \texttt{User32.dll} for windowing related aspects \cite{UCDispatcher}, meaning that the retrieved function address must belong to \texttt{User32.dll}'s memory address range. Such range is determined by retrieving an handle for \texttt{User32.dll} through \texttt{GetModuleHandle}, building an \texttt{IMG} object from it utilizing Pin's \texttt{IMG\_FindByAddress} and invoking the usual \texttt{IMG\_LowAddress} and \texttt{IMG\_HighAddress} functions. 
\item The name of the callback is determined by creating a \texttt{RTN} object from the function address via \texttt{RTN\_FindByAddress} and consequently invoking \texttt{RTN\_Name} on such object.
\item The callback occurrence and the callback name is recorded in the log.
\end{enumerate}

### 2018-07-09 introduction.tex
First Batch of fixes (Emilio's)

\chapter{Introduction}

@@

The heterogeneity of Windows libraries used in malware and the lack of well-structured documentation for their prototypes and input/output parameters posed a significant implementation challenge, which BlueTracer addresses by leveling out and integrating reliable external sources (two industry projects: Dr. Memory and CISCO PyREBox) that account for more than 17,000 existing function prototypes. An important engineering effort was also required to make best use of Intel Pin’s capabilities with respect to run-time CPU and memory costs, as well as to address some inherent limitations that emerged on the field.

@@

\paragraph{Structure of the Thesis.}
The remaining part of this thesis is structured as follows.
Chapter 1 describes the major \textit{API hooking} techniques present in literature, outlining their stengths and weaknesses, especially from a detection point of view. It also introduces the concept of Dynamic Binary Instrumentation (DBI), which constitutes BlueTracer's core.
Chapter 3 focuses on the implementation of the tool, on its structure and the design choices which were made during its development.
Chapter 4 illustrates the experimental results and assesses the tool's effectiveness.
Finally, we present concluding remarks in Chapter 5, together with possible future developments.

>>>>

\chapter{Introduction} \label{Intro}

@@

The heterogeneity of Windows libraries used in malware and the lack of well-structured documentation for their prototypes and input/output parameters posed a significant implementation challenge, which BlueTracer addresses by leveling out and integrating reliable external sources (two industry projects: Dr. Memory \cite{DrMemory} and CISCO PyREBox \cite{Pyrebox}) that account for more than 17,000 existing function prototypes. An important engineering effort was also required to make best use of Intel Pin’s capabilities with respect to run-time CPU and memory costs, as well as to address some inherent limitations that emerged on the field.

@@

\paragraph{Structure of the Thesis.}
The remaining part of this thesis is structured as follows.
Chapter \ref{background} describes the major \textit{API hooking} techniques present in literature, outlining their stengths and weaknesses, especially from a detection point of view. It also introduces the concept of Dynamic Binary Instrumentation (DBI), which constitutes BlueTracer's core.
Chapter \ref{implementation} focuses on the implementation of the tool, on its structure and the design choices which were made during its development.
Chapter \ref{conclusion} illustrates the experimental results and assesses the tool's effectiveness.
Finally, we present concluding remarks in Chapter 5, together with possible future developments.
